An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-29 uai2017/paper.tex

This model is adapted from~\citep{langevin_2011} by increasing the number of
samples to 1 million.  The parameters are $\theta = \langle \theta_1,\theta_2
\rangle$, and the generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i | \theta)^{1/K}$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $K=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

Figure~\ref{fig:gauss_mix_2} shows that {\sc MHminibatch} dominates in terms of
speed and efficiency. The histograms of the (final) minibatch sizes used each
iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 172, more than an order of
magnitude better compared to the other two methods (averages are 12562 and
67508). \textbf{Daniel: Xinlei, these numbers don't match
Figure~\ref{fig:gauss_mix_2} and the figures need to be updated so the scatter
plot has the correct method names and the scaling key, while the histograms have
the correct names and the seaborn background.} We
further ran 10 runs of mixture of Gaussians experiments, and the average and
standard deviation values of the mean minibatch size for {\sc MHminibatch}, {\sc
AustereMH(c)}, {\sc MHSubLhd} are listed in table ~\ref{tab:gaussianres}. Sizes
correspond to the running times of the methods, excluding the likelihood
computation of all data points for {\sc MHSubLhd}, which would drastically
increase running time.

\begin{table}[t]
\caption{Average minibatch sizes ($\pm$ one standard deviation) on the Gaussian
mixture model. The averages are taken over 10 independent trials (3000 samples
each). \textbf{Daniel: Xinlei, can you add in one more decimal point for all the
numbers in this table, like I do in the Logistic Regression table?}}
% Daniel: this `small` command was carried over from the LR table which couldn't
% fit in a column. It's not needed here.
%\small 
\label{tab:gaussianres}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method} & \textbf{Average of MB Sizes} \\
\hline \\
{\sc MHminibatch}   & $182.3\pm 11.4$ \\
{\sc AustereMH(c)}  & $13540.5\pm 1521.4$ \\
{\sc MHSubLhd}      & $65758.9\pm 3222.6$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{table}[t]
\caption{Gaussian Mixture Model Statistics \textbf{TODO Xinlei, please fix this
table.}}
\small
\label{tab:poissons}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Metric} & {\sc MHminibatch} & {\sc AustereMH} & {\sc MHSubLhd} \\
\hline
\T Equation~\ref{eq:log_prob_poissons} & $-1306.2 \pm 20.7$ & $-1619.6 \pm 25.3$ & $-1301.5 \pm 44.4$ \\
Chi-Squared & $5706.8 \pm 2271.5$ & $ 10501.3 \pm 2076.5$ & $ 3711.7 \pm 1182.4$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\textbf{Daniel: Xinlei, you should be in charge of checking this section and
adding some stuff to Table~\ref{tab:poissons}, perhaps averages plus or minus
one standard deviation.}

@@

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\log p(c_1, \ldots, c_{N_{\rm bins}} | P_1, \ldots, P_{N_{\rm bins}}) =
\sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)).
\end{equation}
Table~\ref{tab:poissons} shows the likelihoods. To facilitate interpretation we
perform significance tests using Chi-Squared distribution (also in
Table~\ref{tab:poissons}). The table provides the mean likelihood value and mean 
Chi-Squared test statistics value as well as their standard deviations.  Our likelihood values 
lies between~\citep{cutting_mh_2014} and~\citep{icml2014c1_bardenet14}, 
but the variance of their likelihood values are high and
ordering changes depending on the range of samples generated. It should be noted 
that we are not aiming at optimizing the likelihood values nor the Chi-Squared statistics.
We use these values to show the extent of correctness. \textbf{Daniel:
Xinlei, one reviewer had a problem with this, and I agree. The easiest fix here
is to simply report standard deviation values. Also, we should emphasize that we
are not aiming to get the best score by this metric but simply something that
exhibits some form of ``correctness.''}

>>>>

This model is adapted from~\citep{langevin_2011} by increasing the number of
samples to 1 million.  The parameters are $\theta = \langle \theta_1,\theta_2
\rangle$, and the generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i | \theta)^{1/K}$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $K=10,000$.  Taking logs, we get the target as shown
in the far right of Figure~\ref{fig:gauss_mix_1}.

@@

Figure~\ref{fig:gauss_mix_2} shows that {\sc MHminibatch} dominates in terms of
speed and efficiency. The histograms of the (final) minibatch sizes used each
iteration show that our method consumes significantly less data; the
distribution is short-tailed and the mean is 172, more than an order of
magnitude better compared to the other two methods (averages are 12562 and
67508). We further ran 10 runs of mixture of Gaussians experiments and report
minibatch sizes in Table~\ref{tab:gaussianres}. Sizes correspond to the running
times of the methods, excluding the likelihood computation of all data points
for {\sc AustereMH(nc)} and {\sc MHSubLhd}, which would drastically increase
    running time.

\begin{table}[t]
\caption{Average minibatch sizes ($\pm$ one standard deviation) on the Gaussian
mixture model. The averages are taken over 10 independent trials (3000 samples
each).}
% Daniel: this `small` command was carried over from the LR table which couldn't
% fit in a column. It's not needed here.
%\small 
\label{tab:gaussianres}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Method} & \textbf{Average of MB Sizes} \\
\hline \\
{\sc MHminibatch}   & $182.3\pm 11.4$ \\
{\sc AustereMH(c)}  & $13540.5\pm 1521.4$ \\
{\sc MHSubLhd}      & $65758.9\pm 3222.6$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

\begin{table}[t]
\caption{Gaussian Mixture Model Statistics}
\small
\label{tab:poissons}
\vskip 0.15in
\begin{center}
\begin{tabular}{l l l l}
\textbf{Metric/Method} & {\sc MHminibatch} & {\sc AustereMH(c)} & {\sc MHSubLhd} \\
\hline
\T Equation~\ref{eq:log_prob_poissons} & $-1306.2 \pm 20.7$ & $-1619.6 \pm 25.3$ & $-1301.5 \pm 44.4$ \\
Chi-Squared & $5706.8 \pm 2271.5$ & $ 10501.3 \pm 2076.5$ & $ 3711.7 \pm 1182.4$ \\
\end{tabular}
\end{center}
\vskip -0.1in
\end{table}

@@

@@

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\log p(c_1, \ldots, c_{N_{\rm bins}} | P_1, \ldots, P_{N_{\rm bins}}) =
\sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)).
\end{equation}
Table~\ref{tab:poissons} shows the likelihoods. To facilitate interpretation we
perform significance tests using Chi-Squared distribution (also in
Table~\ref{tab:poissons}). The table provides the mean likelihood value and mean
Chi-Squared test statistics value as well as their standard deviations.  Our
likelihood values lies between~\citep{cutting_mh_2014}
and~\citep{icml2014c1_bardenet14}, but we note that we are not aiming to
optimize the likelihood values or the Chi-Squared statistics.  We use these
values to show the extent of correctness.
